Medical Imaging Classification with Cold-Atom Reservoir Computing using Auto-Encoders and Surrogate-Driven Training\\

### 2025-06-01 paper/main.tex
Updates from Overleaf

In QRC, classical input data \( u(t) \) is encoded into quantum states \( |\psi(t)\rangle \),
which evolve under a fixed Hamiltonian \( H \):
\begin{equation}
    \ket{\psi(t+\Delta t)}
    = U \let{\psi(t)}
    = e^{-iH\Delta t} \ket{\psi(t)},
\end{equation}
where \( U \) is the unitary evolution operator and \(\Delta t\) is the difference between consecutive timestamps. Measurements of observables \( \hat{O} \) yield the outputs
\begin{equation}
    y(t) = \langle \psi(t) | \hat{O} | \psi(t) \rangle.
\end{equation}
The output weights are trained classically, while the quantum reservoir remains fixed.

@@

%============================================
% METHODOLOGY
%============================================
\section{Methodology}
\subsection{Proposed System Architecture}
Our proposed pipeline is a Quantum-Guided Autoencoder with a 
Reservoir Surrogate (QGARS). It is a hybrid architecture that 
synergizes classical autoencoding techniques with quantum reservoir processing to achieve efficient learning of features. These are expected to better match with the reservoir computing layer, whose embeddings are used to perform a classification task.

@@

    \item \textbf{Quantum Reservoir Layer:}
        \begin{itemize}
            \item \textbf{Parameter Mapping:}
            \(g(\bm{z})\) maps the latent representation \(\bm{z}\) to the local detuning frequencies
            \(\Delta_{\mathrm{i}}\) of the Rydberg Atoms.
            \item \textbf{Quantum Evolution:}
            Evolves the system under a Hamiltonian \(H(\bm{\Delta)}\), resulting in a quantum 
            state \(|\psi(\bm{\Delta})\rangle\).
            \item \textbf{Measurement:}
            Performs measurements on the quantum state to obtain embeddings \(\bm{q} \in \mathbb{R}^k\), which serve as training targets for the Surrogate Model or as 
            inputs for the classification task later on.
        \end{itemize}

@@

    \item \textbf{Linear Classifier:}
    A classical neural network that maps the quantum embeddings \(\bm{q}\) to 
    predicted class labels \(\hat{y}\).

@@

If we measure all of the $3$ proposed observables, the total count of the values measured at a single time step $O_N$ is the sum of the counts for each type of observable:
\begin{equation}
    \begin{aligned}
    O_N &= N + \binom{N}{2} + \binom{N}{3}.
    \end{aligned}
    \label{eq:embeddings_per_timestep}
\end{equation}
This set of $O_N$ measurements is repeated at $T$ successive time steps during the evolution of the quantum reservoir. Consequently, the total dimensionality, $\lvert \bm q \rvert$, of the final QRC embedding vector is:
\begin{equation}
    \lvert \bm q \rvert = T O_N.
    \label{eq:total_embedding_dimension}
\end{equation}
This formulation shows that the embedding dimension scales polynomially with the number of qubits $N$ and the number of timesteps $T$ as $\mathcal{O}(TN^3)$ for the considered observables.

@@

\begin{itemize}
    \item A standard autoencoder architecture.
    \item A linear classifier that maps quantum embeddings to class labels.
    \item A classification loss computed on quantum embeddings.
    \item A weighting parameter $\lambda$ to control the trade-off between reconstruction and classification objectives~\ref{eq:composed_loss_function}.
    \item A differentiable surrogate model that approximates the quantum reservoir's behavior, enabling end-to-end backpropagation through the entire pipeline. The frequency of quantum updates is controlled by a hyperparameter, allowing for flexible integration of quantum dynamics into the training process.
\end{itemize}

@@

    \item \textbf{Evolution Time:}
    The time we let the quantum reservoir evolve under the Hamiltonian~\ref{eq:rydberg_hamiltonian}.
\end{itemize}

@@

We can use Eq.~\ref{eq:total_embedding_dimension} to determine the dimension of our quantum embeddings \(\lvert \bm q \rvert \) when using the mentioned parameters:

@@

\subsection{Comparison Methods}
We benchmarked our model against the following classical methods:
\begin{itemize}
    \item \textbf{PCA/Autoencoder + Linear Classifier:} 
    A linear classifier trained on the PCA-reduced features or the latent representations from the autoencoder. Essentially, this model is the same as the one used to classify the quantum embeddings described in section~\ref{linear_classifier_for_the_qrc_embeddings}, but it only receives the classically extracted features as input.

@@

We also benchmarked our QGARS encoding strategy
against the performance of the Linear Mapping of
QRC embeddings extracted from the latent representation 
of data using other feature reduction methods, such as
PCA and the Classical Autoencoder.

@@

\subsection{Parameter Sweep Strategy}
To optimize the performance of our quantum-guided autoencoder, 
we conducted a systematic parameter sweep over the following hyperparameters:
\begin{itemize}
    \item \textbf{Guided Lambda Parameter (\( \lambda \))}: 
    To analyse the effect that the trade-off between reconstruction and classification objectives in the loss function~\ref{eq:composed_loss_function} affects
    the QRC

@@

\subsection{Dimensionality Reduction Method Comparison}
We first compare classification accuracy of the Linear Classifier on the QRC embeddings across the different dimensionality reduction methods PCA, vanilla autoencoder, and our QGARS pipeline on each dataset. Table~\ref{tab:qrc_reduction_accuracy} summarizes mean accuracy and standard deviation over five runs, using 12 features/qubits. These results show that the QGARS pipeline is able to outperform traditional feature reduction methods commonly used as preprocessing steps on QRC pipelines.

\begin{table}[!b]
  \caption{Classification accuracy (\%) on QRC with different dimensionality reduction methods (2000 training images and 400 testing images)}
  \label{tab:qrc_reduction_accuracy}
  \centering
  \begin{tabular}{lccc}
    \hline
    Method         & Synthetic  & CVC-ClinicDB & MNIST(0/1) \\
    \hline
    PCA + QRC      &  00 ± 0.0  &  84.95 ± 2.44    &  97.70 ± 0.89 \\
    AE + QRC       &  00 ± 0.0  &  73.75 ± 1.91    &  87.05 ± 1.85 \\
    \textbf{QGARS + QRC}    &  \textbf{00 ± 0.0} &  \textbf{88.90 ± 1.47} &  \textbf{00 ± 0.0} \\
    \hline
  \end{tabular}
\end{table}

\begin{figure}[!t]
    \centering
    \includegraphics[width=\linewidth]{images/results/generated_polyp_dataset/n_qubit_qrc_performance_by_reduction_method.pdf} 
    \caption{Accuracy of the different dimensionality reduction methods for QRC as a function of the number of qubits with the Synthetic Polyp Dataset. (green) QGARS, (blue) PCA and (red) Autoencoder. The dashed lines represent the training accuracy, while the whole lines are the test accuracies.}
    \label{fig:n_qubits_accuracy}
\end{figure}

\begin{figure*}[!ht]
  \centering
  \includegraphics[width=0.32\textwidth]{images/results/generated_polyp_dataset/quantum_embeddings_2d_tsne_PCA.pdf}
  \includegraphics[width=0.32\textwidth]{images/results/generated_polyp_dataset/quantum_embeddings_2d_tsne_autoencoder.pdf}
  \includegraphics[width=0.32\textwidth]{images/results/generated_polyp_dataset/quantum_embeddings_2d_tsne_guided_autoencoder.pdf}
  \caption{2-dimensional t-SNE projections of the 4768-dimensional QRC embeddings for (left) PCA + QRC, (center) AE + QRC, (right) QGARS + QRC}
  \label{fig:embeddings_tsne}
\end{figure*}

@@

In order to have a visual demonstration of how well the quantum reservoir embeddings separate classes, we project the high-dimensional embeddings down to $3$ dimensions via t-distributed Stochastic Neighbour Embedding (t-SNE). Fig.~\ref{fig:embeddings_tsne} shows side-by-side plots for PCA → QRC, AE → QRC, and QGARS → QRC pipelines. The QGARS embeddings exhibit the tightest and most distinct clusters, indicating superior discriminative structure.

@@

%\subsubsection{Impact of Guided Lambda Parameter}
We verified that varying the guidance weighting $\lambda$ in Eq.~\ref{eq:composed_loss_function} has significant impact on the performance of our accuracy. To assess its influence, Fig.~\ref{fig:lambda_sweep} plots test accuracy vs.\ $\lambda$.

@@

\begin{table}[tb]
  \caption{Classification accuracy (\%) for classical baselines (2000 training images and 400 testing images)}
  \label{tab:classical_comp}
  \centering
  \begin{tabular}{lccc}
    \hline
    Method            & Synthetic & CVC-ClinicDB & MNIST (0/1) \\
    \hline
    PCA + Linear      &  82.0 ± 0.79    &  71.05 ± 1.35   &  99.90 ± 0.12    \\
    AE + Linear       &  80.3 ± 3.01    &  68.25 ± 2.31   &  98.80 ± 0.29    \\
    PCA + NN          &  85.85 ± 1.24   &  91.65 ± 1.38   &  99.80 ± 0.19    \\
    AE + NN           &  83.75 ± 2.21   &  90.50 ± 1.72   &  99.90 ± 0.12    \\
    \hline
  \end{tabular}
\end{table}

\noindent
In contrast, the QGARS pipeline achieves a maximum of $\mathbf{0.0\%}$ and $\mathbf{90.50\%}$, $\mathbf{0.0\%}$ on the same datasets (see Table~\ref{tab:qrc_reduction_accuracy}).

@@

    \item \textbf{Experiment on Real Quantum Hardware:}
    Although the experiments we conducted on simulators showed promising results, to validate the usefulness of our pipeline in the NISQ era, we still need to run experiences on real QPUs with more qubits.

@@

We have presented Quantum-Guided Autoencoder with Reservoir Surrogate (QGARS), a hybrid quantum-classical architecture that integrates a classical autoencoder with a neutral-atom quantum reservoir and a differentiable surrogate model to enable end‐to‐end training. By jointly optimizing reconstruction and classification losses, QGARS learns compact and discriminative features tailored for quantum reservoir computing. Our experiments on synthetic polyp, CVC-ClinicDB, and binary MNIST datasets demonstrate that QGARS outperforms classical PCA‐ and autoencoder‐based baselines that are commonly used for Quantum Reservoir Computing Pipelines. Ablation studies highlight the importance of guidance weight \(\lambda\), surrogate update frequency, and reservoir parameters. Although challenges remain—particularly regarding high‐dimensional embeddings, surrogate fidelity, and classical overhead, our proposed extensions and hardware considerations chart a clear path toward robust, scalable quantum‐enhanced medical imaging pipelines in the NISQ era.

>>>>

In QRC, classical input data \( u(t) \) is encoded into quantum states \( |\psi(t)\rangle \),
which evolve under a fixed Hamiltonian \( H \):
\begin{equation}
    \ket{\psi(t+\Delta t)}
    = U \ket{\psi(t)}
    = e^{-iH\Delta t} \ket{\psi(t)},
\end{equation}
where \( U \) is the unitary evolution operator and \(\Delta t\) is the difference between consecutive timestamps. Measurements of observables \( \hat{O} \) yield the outputs
\begin{equation}
    y(t) = \langle \psi(t) | \hat{O} | \psi(t) \rangle.
\end{equation}
The output weights are trained classically, while the quantum reservoir remains fixed.

@@

%============================================
% METHODOLOGY
%============================================
\section{Methodology}
\subsection{Proposed System Architecture}
Our proposed pipeline is a Quantum-Guided Autoencoder with a 
Reservoir Surrogate (QGARS). It is an hybrid architecture that 
synergizes classical autoencoding techniques with quantum reservoir processing to achieve efficient learning of features. These are expected to better match with the reservoir computing layer, whose embeddings are used to perform a classification task.

@@

    \item \textbf{Quantum Reservoir Layer:}
        \begin{itemize}
            \item \textbf{Parameter Mapping:}
            \(g(\bm{z})\) maps the latent representation \(\bm{z}\) to the local detuning frequencies
            \(\Delta_{\mathrm{i}}\) of the Rydberg Atoms.
            \item \textbf{Quantum Evolution:}
            Evolves the system under a Hamiltonian \(H(\bm{\Delta)}\), resulting in a quantum 
            state \(|\psi(\bm{\Delta})\rangle\).
            \item \textbf{Measurement:}
            Performs measurements on the quantum state to obtain embeddings \(\bm{q} \in \mathbb{R}^k\), which serve as training targets for the surrogate model or as 
            inputs for the classification task later on.
        \end{itemize}

@@

    \item \textbf{Linear Classifier:}
    A classical neural network that maps the quantum embeddings \(\bm{q}\) to 
    predicted class labels \(\bm{\hat{y}}\).

@@

If we measure all of the three proposed observables, the total count of the values measured at a single time step $O_N$ is the sum of the counts for each type of observable:
\begin{equation}
    \begin{aligned}
    O_N &= N + \binom{N}{2} + \binom{N}{3}.
    \end{aligned}
    \label{eq:embeddings_per_timestep}
\end{equation}
This set of $O_N$ measurements is repeated at $T$ successive time steps during the evolution of the quantum reservoir. Consequently, the total dimensionality, $\lvert \bm q \rvert$, of the final QRC embedding vector is:
\begin{equation}
    \lvert \bm q \rvert = T O_N.
    \label{eq:total_embedding_dimension}
\end{equation}
This formulation shows that the embedding dimension scales polynomially with the number of qubits $N$ and the number of timesteps $T$ as $\mathcal{O}(TN^3)$ for the considered observables.

@@

\begin{itemize}
    \item A standard autoencoder architecture.
    \item A linear classifier that maps quantum embeddings to class labels.
    \item A classification loss computed on quantum embeddings.
    \item A weighting parameter $\lambda$ to control the trade-off between reconstruction and classification objectives~\eqref{eq:composed_loss_function}.
    \item A differentiable surrogate model that approximates the quantum reservoir's behavior, enabling end-to-end backpropagation through the entire pipeline. The frequency of quantum updates is controlled by a hyperparameter, allowing for flexible integration of quantum dynamics into the training process.
\end{itemize}

@@

    \item \textbf{Evolution Time:}
    The time we let the quantum reservoir evolve under the Hamiltonian~\eqref{eq:rydberg_hamiltonian}.
\end{itemize}

@@

We can use~\eqref{eq:total_embedding_dimension} to determine the dimension of our quantum embeddings \(\lvert \bm q \rvert \) when using the mentioned parameters:

@@

\subsection{Comparison Methods}
We benchmarked the prediction accuracy of our model against the following classical methods:
\begin{itemize}
    \item \textbf{PCA/Autoencoder + Linear Classifier:} 
    A linear classifier trained on the PCA-reduced features or the latent representations from the autoencoder. Essentially, this model is the same as the one used to classify the quantum embeddings described in section~\ref{linear_classifier_for_the_qrc_embeddings}, but it only receives the classically extracted features as input.

@@

We also benchmarked our QGARS encoding strategy
against the performance of the Linear Mapping of
QRC embeddings extracted from the latent representation 
of data using other feature reduction methods, such as
PCA and the classical autoencoder.

The accuracy results were measured with 2000 training images and 400 testing images over 5 different seeds.

@@

\subsection{Parameter Sweep Strategy}
To optimize the performance of our quantum-guided autoencoder, 
we conducted a systematic parameter sweep over the following hyperparameters:
\begin{itemize}
    \item \textbf{Guided Lambda Parameter (\( \lambda \))}: 
    To analyse the effect that the trade-off between reconstruction and classification objectives in the loss function~\eqref{eq:composed_loss_function} affects
    the QRC

@@

\subsection{Dimensionality Reduction Method Comparison}
We first compare classification accuracy of the linear classifier on the QRC embeddings across the different dimensionality reduction methods PCA, vanilla autoencoder, and our QGARS pipeline on each dataset. Table~\ref{tab:qrc_reduction_accuracy} summarizes mean accuracy and standard deviation over five runs, using 12 features/qubits. These results show that the QGARS pipeline is able to outperform traditional feature reduction methods commonly used as preprocessing steps on QRC pipelines.

\begin{table}[!b]
  \caption{Mean ± standard deviation of the classification accuracy (\%) QRC with different dimensionality reduction methods}
  \centering
  \label{tab:qrc_reduction_accuracy}
  \begin{tabular}{lccc}
    \hline
    Method         & Synthetic  & CVC-ClinicDB & MNIST(0/1) \\
    \hline
    PCA + QRC      &  79.20 ± 2.53  &  84.95 ± 2.31   &  97.70 ± 0.97 \\
    AE + QRC       &  77.50 ± 2.70  & 73.75 ± 1.90   &  87.05 ± 2.09 \\
    \textbf{QGARS + QRC}    &  \textbf{89.45 ± 2.06} &  \textbf{88.90 ± 1.65} &  \textbf{97.00 ± 2.37} \\
    \hline
  \end{tabular}
\end{table}

\begin{figure}[!t]
    \centering
    \includegraphics[width=\linewidth]{images/results/generated_polyp_dataset/n_qubit_qrc_performance_by_reduction_method.pdf} 
    \caption{Accuracy of the different dimensionality reduction methods for QRC as a function of the number of qubits with the synthetic polyp dataset. (green) QGARS, (blue) PCA and (red) Autoencoder. The dashed lines represent the training accuracy, while the whole lines are the test accuracies.}
    \label{fig:n_qubits_accuracy}
\end{figure}

\begin{figure*}[!ht]
  \centering
  \includegraphics[width=0.32\textwidth]{images/results/generated_polyp_dataset/quantum_embeddings_2d_tsne_PCA.pdf}
  \includegraphics[width=0.32\textwidth]{images/results/generated_polyp_dataset/quantum_embeddings_2d_tsne_autoencoder.pdf}
  \includegraphics[width=0.32\textwidth]{images/results/generated_polyp_dataset/quantum_embeddings_2d_tsne_guided_autoencoder.pdf}
  \caption{two-dimensional t-SNE projections of the 4768-dimensional QRC embeddings for (left) PCA + QRC, (center) AE + QRC, (right) QGARS + QRC}
  \label{fig:embeddings_tsne}
\end{figure*}

@@

In order to have a visual demonstration of how well the quantum reservoir embeddings separate classes, we project the high-dimensional embeddings down to three dimensions via t-distributed Stochastic Neighbour Embedding (t-SNE). Fig.~\ref{fig:embeddings_tsne} shows side-by-side plots for PCA → QRC, AE → QRC, and QGARS → QRC pipelines. The QGARS embeddings exhibit the tightest and most distinct clusters, indicating superior discriminative structure.

@@

%\subsubsection{Impact of Guided Lambda Parameter}
We verified that varying the guidance weighting $\lambda$ in~\eqref{eq:composed_loss_function} has significant impact on the performance of our accuracy. To assess its influence, Fig.~\ref{fig:lambda_sweep} plots test accuracy vs.\ $\lambda$.

@@

\begin{table}[tb]
  \caption{Classification accuracy (\%) for classical baselines (2000 training images and 400 testing images)}
  \label{tab:classical_comp}
  \centering
  \begin{tabular}{lccc}
    \hline
    Method            & Synthetic & CVC-ClinicDB & MNIST (0/1) \\
    \hline
    PCA + Linear      &  82.00 ± 0.88    &  71.05 ± 1.51   &  99.90 ± 0.14   \\
    AE + Linear       &  80.30 ± 3.37    &  68.25 ± 2.58   &  98.80 ± 0.33   \\
    PCA + NN          &  85.85 ± 1.39   &  91.65 ± 1.55   &  99.80 ± 0.21    \\
    AE + NN           &  83.75 ± 2.47    &  90.50 ± 1.93   &  99.90 ± 0.14    \\
    \hline
  \end{tabular}
\end{table}

\noindent
In contrast, the QGARS pipeline achieves a maximum of $\mathbf{92.0\%}$ and $\mathbf{90.5\%}$, $\mathbf{99.0\%}$ on the same datasets (see Table~\ref{tab:qrc_reduction_accuracy}).

@@

    \item \textbf{Experiment on Real Quantum Hardware:}
    Although the experiments we conducted on simulators showed promising results, to validate the usefulness of our pipeline in the NISQ era, we still need to run experiences on real quantum processing units (QPUs) with more qubits.

@@

We have presented Quantum-Guided Autoencoder with Reservoir Surrogate (QGARS), a hybrid quantum-classical architecture that integrates a classical autoencoder with a neutral-atom quantum reservoir and a differentiable surrogate model to enable end‐to‐end training. By jointly optimizing reconstruction and classification losses, QGARS learns compact and discriminative features tailored for quantum reservoir computing. Our experiments on synthetic polyp, CVC-ClinicDB, and binary MNIST datasets demonstrate that QGARS outperforms classical PCA‐ and autoencoder‐based baselines that are commonly used for quantum reservoir computing pipelines. Ablation studies highlight the importance of guidance weight \(\lambda\), surrogate update frequency, and reservoir parameters. Although challenges remain—particularly regarding high‐dimensional embeddings, surrogate fidelity, and classical overhead, our proposed extensions and hardware considerations chart a clear path toward robust, scalable quantum‐enhanced medical imaging pipelines in the NISQ era.
